---
title: "Statistical Analysis"
output:
  html_document:
  code_folding: hide
  toc: true
  toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(viridis)
library(rvest)
library(purrr)
library(broom)
library(modelr)
library(mgcv)
library(patchwork)
library(plotly)
library(maps)
library(leaflet)
library(rgdal)
library(maptools)
library(BAMMtools)
library(spdep)
library(kableExtra)
```

###Importing and cleaning EMS data in NYC: 
```{r warning=FALSE}
ems_data_clean = read.csv("./data/EMS_Incident_Dispatch_Data.csv") %>%
  select(INITIAL_SEVERITY_LEVEL_CODE, FINAL_SEVERITY_LEVEL_CODE, INITIAL_CALL_TYPE,
         DISPATCH_RESPONSE_SECONDS_QY, INCIDENT_TRAVEL_TM_SECONDS_QY, HELD_INDICATOR, BOROUGH,
         ZIPCODE, INCIDENT_DISPOSITION_CODE, INCIDENT_DATETIME) %>% 
  janitor::clean_names() %>% 
  separate(col = incident_datetime, into = c('date', 'time'), sep = ' ') %>% 
  separate(col = date, into = c("month","day"), sep = '/') %>% 
  mutate(month = factor(month, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"), labels = c("Jan", "Feb", "Mar", "Apr", "May", "June", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"), ordered = TRUE)) %>% 
  mutate(arrival_outcome = ifelse(incident_disposition_code == "83", "dead", "alive"),
         arrival_outcome = recode_factor(arrival_outcome, `0` = "alive", `1` = "dead"),
         initial_severity_level_code = factor(initial_severity_level_code, 
 levels = c("1", "2","3", "4", "5", "6", "7", "8", "9"), ordered = TRUE),
         final_severity_level_code = factor(final_severity_level_code, 
                                            levels = c("1", "2", "3", "4", "5", "6", "7", "8"), ordered = TRUE), 
         held_indicator = recode(held_indicator, "N" = "no", "Y" = "yes")) %>%
  mutate(neighbourhood = recode(zipcode, "10026" = "central harlem", "10027" = "central harlem", "10030" = "central harlem", "10037" = "central harlem", "10039" = "central harlem", "10001" = "chelsea and clinton", "10001" = "chelsea and clinton", "10011" = "chelsea and clinton", "10018" = "chelsea and clinton", "10019" = "chelsea and clinton", "10020" = "chelsea and clinton", "10036" = "chelsea and clinton",  "10029" = "east harlem", "10035" = "east harlem", "10010" = "gramercy park and murray hill", "10016" = "gramercy park and murray hill", "10017" = "gramercy park and murray hill", "10022" = "gramercy park and murray hill", "10012" = "greenwich village and soho", "10013" = "greenwich village and soho", "10014" = "greenwich village and soho", "10004" = "lower manhattan", "10005" = "lower manhattan", "10006" = "lower manhattan", "10007" = "lower manhattan", "10038" = "lower manhattan", "10280" = "lower manhattan", "10002" = "lower east side", "10003" = "lower east side", "10009" = "lower east side", "10021" = "upper east side", "10028" = "upper east side", "10044" = "upper east side", "10065" = "upper east side", "10075" = "upper east side", "10128" = "upper east side", "10023" = "upper west side", "10024" = "upper west side", "10025" = "upper west side", "10031" = "inwood and washington heights", "10032" = "inwood and washington heights", "10033" = "inwood and washington heights", "10034" = "inwood and washington heights", "10040" = "inwood and washington heights" )
  ) %>%  
  drop_na(neighbourhood) %>% 
    select(-incident_disposition_code) 
```

###t-test and ANOVA:
```{r warning=FALSE}
t.test(ems_data_clean$incident_travel_tm_seconds_qy/60 ~ ems_data_clean$arrival_outcome, na.rm = FALSE)

res.aov = aov(incident_travel_tm_seconds_qy/60 ~ neighbourhood, data = ems_data_clean)
# Summary of the analysis
summary(res.aov) 
```
t-test: 
- for those alive: average incident response time = 8.83 minutes 
- for those dead: average incident response time = 4.91 minutes
Since the p-value < 

Anova: 
Since the p-value is less than 0.05, we know there are statistically significant differneces between the mean incident travel time in seconds between neighborhoods.


###Logistic Regression Model and Results Table: 
```{r include= FALSE, warning=FALSE}
ems_data_clean %>%  
    mutate(
    initial_severity_level_code = fct_relevel(initial_severity_level_code, "1")) %>%
    select(incident_travel_tm_seconds_qy, initial_severity_level_code, neighbourhood)
```

```{r}
model = glm(arrival_outcome ~ incident_travel_tm_seconds_qy + initial_severity_level_code + neighbourhood, data = ems_data_clean, family = binomial())

model %>% 
  broom::tidy() %>% 
  mutate(OR = exp(estimate)) %>%
  select(term, log_OR = estimate, OR, p.value) %>% 
  knitr::kable(digits = 3)
```

We demonstrate the odds of death on arrival for each of the variables used in the analysis. 







