---
title: "Exploratory Analysis"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(viridis)
library(rvest)
library(purrr)
library(broom)
library(modelr)
library(mgcv)
library(patchwork)
library(plotly)
library(maps)
library(leaflet)
library(rgdal)
library(maptools)
library(BAMMtools)
library(spdep)
library(kableExtra)
```


```{r warning=FALSE, echo=FALSE}
###Importing and cleaning EMS data in NYC:

ems_data_clean = read.csv("./data/EMS_Incident_Dispatch_Data.csv") %>%
  select(INITIAL_SEVERITY_LEVEL_CODE, FINAL_SEVERITY_LEVEL_CODE, INITIAL_CALL_TYPE,
         DISPATCH_RESPONSE_SECONDS_QY, INCIDENT_TRAVEL_TM_SECONDS_QY, HELD_INDICATOR, BOROUGH,
         ZIPCODE, INCIDENT_DISPOSITION_CODE, INCIDENT_DATETIME) %>% 
  janitor::clean_names() %>% 
  separate(col = incident_datetime, into = c('date', 'time'), sep = ' ') %>% 
  separate(col = date, into = c("month","day"), sep = '/') %>% 
  mutate(month = factor(month, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"), labels = c("Jan", "Feb", "Mar", "Apr", "May", "June", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"), ordered = TRUE)) %>% 
  mutate(arrival_outcome = ifelse(incident_disposition_code == "83", "dead", "alive"),
         arrival_outcome = recode_factor(arrival_outcome, `0` = "alive", `1` = "dead"),
         initial_severity_level_code = factor(initial_severity_level_code, 
 levels = c("1", "2","3", "4", "5", "6", "7", "8", "9"), ordered = TRUE),
         final_severity_level_code = factor(final_severity_level_code, 
                                            levels = c("1", "2", "3", "4", "5", "6", "7", "8"), ordered = TRUE), 
         held_indicator = recode(held_indicator, "N" = "no", "Y" = "yes")) %>%
  mutate(neighbourhood = recode(zipcode, "10026" = "central harlem", "10027" = "central harlem", "10030" = "central harlem", "10037" = "central harlem", "10039" = "central harlem", "10001" = "chelsea and clinton", "10001" = "chelsea and clinton", "10011" = "chelsea and clinton", "10018" = "chelsea and clinton", "10019" = "chelsea and clinton", "10020" = "chelsea and clinton", "10036" = "chelsea and clinton",  "10029" = "east harlem", "10035" = "east harlem", "10010" = "gramercy park and murray hill", "10016" = "gramercy park and murray hill", "10017" = "gramercy park and murray hill", "10022" = "gramercy park and murray hill", "10012" = "greenwich village and soho", "10013" = "greenwich village and soho", "10014" = "greenwich village and soho", "10004" = "lower manhattan", "10005" = "lower manhattan", "10006" = "lower manhattan", "10007" = "lower manhattan", "10038" = "lower manhattan", "10280" = "lower manhattan", "10002" = "lower east side", "10003" = "lower east side", "10009" = "lower east side", "10021" = "upper east side", "10028" = "upper east side", "10044" = "upper east side", "10065" = "upper east side", "10075" = "upper east side", "10128" = "upper east side", "10023" = "upper west side", "10024" = "upper west side", "10025" = "upper west side", "10031" = "inwood and washington heights", "10032" = "inwood and washington heights", "10033" = "inwood and washington heights", "10034" = "inwood and washington heights", "10040" = "inwood and washington heights" )
  ) %>%  
  drop_na(neighbourhood) %>% 
    select(-incident_disposition_code) 
```


#
#

# Death Counts by Month

#

```{r warning=FALSE, echo=FALSE}

ems_data_clean %>% 
  select(arrival_outcome, month) %>% 
  group_by(month, arrival_outcome) %>%
  drop_na(arrival_outcome) %>% 
  summarise_(n = ~n()) %>%
  filter(arrival_outcome == "dead") %>% 
  plot_ly(y = ~n, x = ~month, type = "bar",
          color = ~month, alpha = 0.7) %>% 
  layout(title = "Death Count on Arrival for each Month in 2018",
    xaxis = list(title = "Month"),
         yaxis = list(title = 'Dead on Arrival (n)'),
    showlegend = FALSE)
```

The number of individuals dead upon the arrival of EMS responders peaks in the months of January (over 200 deaths) and June (nearly 200 deaths). Death numbers are the highest in the Winter and Summer months, while September has the lowest death rate.

#
#

# Death Count by Neighborhood:

#

```{r warning=FALSE, echo=FALSE}

ems_data_clean %>% 
  select(arrival_outcome, neighbourhood) %>% 
  group_by(neighbourhood, arrival_outcome) %>%
  drop_na(arrival_outcome) %>% 
  summarise_(n = ~n()) %>%
  filter(arrival_outcome == "dead") %>% 
  plot_ly(y = ~n, x = ~neighbourhood, type = "bar",
          color = ~neighbourhood) %>% 
  layout(title = "Death Count on Arrival by NYC Neighborhood",
    xaxis = list(title = "Neighborhood"),
         yaxis = list(title = 'Dead on Arrival (n)'),
    showlegend = FALSE)
```

The highest death count on arrival was in the Upper West Side, followed by the Lower East Side, Inwood and Washington Heights, and Central Harlem. While the lowest death count on arrival was in Lower Manhattan, Greenwich Village and Soho.

#
#

# Death Count by NYC Neighborhood Over Time, 2018:

#

```{r warning=FALSE, echo=FALSE}

deaths_scatter = ems_data_clean %>%
select(arrival_outcome, neighbourhood, month) %>%
group_by(arrival_outcome, neighbourhood, month) %>% 
drop_na(arrival_outcome) %>% 
summarise_(n_death = ~n()) %>%
filter(arrival_outcome %in% "dead") %>%
plot_ly(x = ~month, y = ~n_death, color = ~neighbourhood,type = "scatter", mode = "lines+markers", alpha = 0.5) %>%
  layout(title = 'Death Count on Arrival by NYC Neighborhood Over Time, 2018',
    xaxis = list(title = 'Month'),
         yaxis = list(title = 'Dead on Arrival (n)'),
         legend = list(x = 0.5, y = 0.95))
deaths_scatter
```

Based on NYC EMS calls, the Upper West Side experienced the highest numbers of deaths per month in 2018 compared to all other NYC neighborhoods, on average. Additionally, the neighborhoods of Lower Manhattan and Greenwich Village and Soho experienced the least numbers of deaths per month in 2018.

#

# Mean Travel Time (s) by Neighborhood: 

#

```{r warning=FALSE, echo=FALSE}
ems_data_clean %>% 
  select(arrival_outcome, incident_travel_tm_seconds_qy) %>% 
  group_by(arrival_outcome) %>% 
  drop_na(arrival_outcome) %>%
  summarize(mean_time = mean(incident_travel_tm_seconds_qy))

time_boro = ems_data_clean %>%
  select(neighbourhood, arrival_outcome, incident_travel_tm_seconds_qy) %>%
  group_by(arrival_outcome, neighbourhood) %>%
  drop_na(arrival_outcome) %>%
  summarize(mean_time = mean(incident_travel_tm_seconds_qy)) %>% 
  plot_ly(x = ~neighbourhood, y = ~mean_time/60, color = ~arrival_outcome, group = ~arrival_outcome, type = "bar", colors = "Set2") %>% 
  layout(title = "Mean Travel Time by NYC Neighborhood",
    xaxis = list(title = "Neighborhood"),
         yaxis = list(title = 'Mean EMS Travel Time (s)'),
    showlegend = FALSE)
time_boro
```

The overall mean response time for those who survived was 8.83 minutes, while the average for those who died was 4.91 minutes. In the graph we can see that East and Central Harlem, as well as the Upper West Side, have above the average travel times for alive arrival. And Inwoode and Washington heights, Central Harlem, and Greenwich Village and Soho have above the mean travel times for dead arrival.

#
#

# Severity Code by Travel Time for Each Neighborhood:

#

```{r warning=FALSE, echo=FALSE}

ems_data_clean %>%
  select(neighbourhood, initial_severity_level_code, incident_travel_tm_seconds_qy) %>%
  group_by(neighbourhood, initial_severity_level_code) %>% 
  summarize(mean_time = mean(incident_travel_tm_seconds_qy)) %>%
  plot_ly(x = ~initial_severity_level_code, y = ~mean_time/60, group = ~neighbourhood, color = ~neighbourhood, type = "scatter", mode = "lines+markers", alpha = 0.5) %>% 
  layout(title = "Mean Travel Time by Severity for each Neighborhood",
    xaxis = list(title = "Severity Code"),
         yaxis = list(title = "Mean Travel Time (s)"),
         legend = list(x = 0.5, y = 0.95))
```
With increasing initial severity level code, the average travel response time taken to reach the incidece increases. There is a slower response time for higher intial severity code, and a decrease in response time for the severity score 8 (highest).The slowest times for highest severity are for East Harlem and Lower Manhattan.
Overall, EMS responders response the slowest to East Harlem, then Central Harlem, and then Inwood and Washington Heights.
