---
title: "**Help! I've Fallen and I Can't Get Up: A Report**"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

*Note: The title of this project is a line from a commercial that has been [memed](https://www.google.com/search?q=life+alert+meme&sxsrf=ACYBGNQqIvW74KMRbdauXsERsqWnXo-X_Q:1575226120265&source=lnms&tbm=isch&sa=X&ved=2ahUKEwiWjdvKjpXmAhVHq1kKHccFBD8Q_AUoAXoECAsQAw&biw=1130&bih=1257) and is often the subject of jokes. It is in no way meant to trivialize the emergencies experienced by 911 callers or the work of EMS responders. 

#

## Motivation and Project Goals

In New York City, emergencies abound. It’s hard to imagine a day in New York without the sound of wailing sirens and flashing emergency vehicle lights. 

It is expected that when it comes to emergency medical services (EMS), faster is better. Most EMS operations aim to have a response time of 8 minute or less for advanced life support (ALS) units responding to life-threatening events.

In 2018, it was estimated that [~46 percent of ambulance responses to serious emergencies took longer than 10 minutes in NYC](https://cbcny.org/research/reviving-ems). The City of New York spends more than [$1.1 billion annually to provide EMS services](https://council.nyc.gov/budget/wp-content/uploads/sites/54/2018/03/FY19-Fire-Department-of-New-York.pdf) to the city, however the money is often not used wisely. Conflicting reports claiming that an 8 minute or less response time may not improve health outcomes call into question whether a response time <8 min is important and if we should we allocate more funds to reduce NYC response times.

This study seeks to assess the association between NYC EMS response time and mortality.


#

## Related Work/Inspiration

Living in New York City and working on the CUMC campus, we recognize the importance of efficient emergency medical services. As aspiring public health officials, we drew inspiration from the Reviving EMS Report issued by the [Citizen’s Budget Commission](link https://cbcny.org/research/reviving-ems). Due to growing concerns of increasing EMS response times to patients residing in lower-income neighborhoods, we wanted to see if increased EMS response time was associated with more deaths upon arrival. With limited EMS budgeting, we hope our project may help city officials better allocate their resources and budgeting to optimize EMS in NYC. 

#

## Initial Questions

Our team wanted to examine the relationship between location in NYC and EMS response time. We wanted to better understand whether it takes EMS responders longer to reach incidents in certain neighborhoods. We also wanted to examine if the length of response time varied by the initial severity code of the call. When an individual calls 911, the incident they are reporting is scored a severity code based on how severe the event is initially deemed to be based on the call. We hypothesize that the travel time would be shortest in response to calls that have the highest initial severity code. 

We focused our research questions on Manhattan, as we live and go to school here and felt that it has a very diverse range of neighborhoods. This also minimizes bias due to distance traffic variables that are difficult to measure.
We posed the following questions:

* Does EMS response travel time vary by neighborhood? 
* Does EMS response travel time vary by whether the person was alive or dead upon arrival? 
* Does EMS response travel time vary by the severity code of the initial call? 
* Do the number of deaths upon arrival of EMS responders vary by month? 
* Do the number of deaths upon arrival of EMS responders vary by NYC neighborhood? 

We wanted to gain stronger insight into these questions because this information would be vital in determining if funding should be designated to improving the EMS response system in NYC. If a faster response time improved your arrival outcome, this could save the lives of many New Yorkers. If response times are not the same across neighborhoods, this would provide us with a better perspective of existing disparities in healthcare and further research could shape interventions to improve emergency responses.


#

## Data
The data source for this project comes from the [NYC EMS Incident Dispatch Data](https://data.cityofnewyork.us/Public-Safety/EMS-Incident-Dispatch-Data/76xm-jjuj), owned by NYCOpenData and provided by the Fire Department of New York City.

The EMS Incident Dispatch Data is generated by the FDNY EMS Computer Aided Dispatch System. The data spans from the time an incident is created in the system to the time that incident is closed. 

For every call/emergency in the system from the start of 2013 to the end of 2018, the database contains information on the following components: 

* Time (initial, closing)

* Call/emergency type (initial, closing, severity)

* Location (borough, dispatch area, zip code, precinct, district (city council, community, school, congressional))

* Response time (incident, dispatch)

As well as some internal quality control measures. 

As noted, to protect personal identifying information in accordance with the Health Insurance Portability and Accountability Act (HIPAA), specific locations of incidents are not included and have been aggregated to a higher level of detail.

Additionally, for the assignment of neighborhood by zip code, we used the [New York State Department of Health ZIP Code Definitions of New York City Neighborhoods](https://www.health.ny.gov/statistics/cancer/registry/appendix/neighborhoods.htm).


#

## Cleaning
The original dataset contains 8.56 million rows of data, spanning across 5 years. 
Due to computing restraints, we narrowed down the dataset before cleaning it in RStudio. We applied the following constraints to the dataset:

* Borough: Manhattan
* Year: 2018
* Valid incident response time (quality control)
* Only non-cancelled calls
* No duplicate calls
* Only calls where EMS was dispatched
* Only calls where the final clinical outcome was not missing or unknown

After we limited the data set, data cleaning steps included separating the date and time variables, recoding relevant variables as factors, creating a clinical outcome variable regarding living status, creating a neighborhood variable based on zip code, and excluding missing values. 


#

## Exploratory analysis

Our exploratory analysis focused on two outcomes of interest: 

* Death upon arrival of EMS Response Team

* Response time

During our exploration, we sought to find any differences in these outcomes depending on neighborhood, and whether these outcomes were independent of each other. 
Initially, we conducted the exploratory analysis using the zip codes originally found in the NYC OpenData dataset, but felt that this was inaccessible and produced hard-to-read graphs. Thus, we used the NYS standardized neighborhoods to make our results more coherent and understandable. 

Additionally, we decided to explore whether these outcomes varied by the time of year, and whether there was a difference in the types of emergencies in each neighborhood. We felt that visual, interactive tools were the easiest way for us to see the data. Below are figures that summarize the inter-relationships of our main variables of focus: 


```{r, include = FALSE, echo = FALSE}
library(tidyverse)
library(patchwork)
library(plotly)
ems_data_clean = read.csv("./data/EMS_Incident_Dispatch_Data.csv") %>%
  select(INITIAL_SEVERITY_LEVEL_CODE, FINAL_SEVERITY_LEVEL_CODE, INITIAL_CALL_TYPE,
         DISPATCH_RESPONSE_SECONDS_QY, INCIDENT_TRAVEL_TM_SECONDS_QY, HELD_INDICATOR, BOROUGH,
         ZIPCODE, INCIDENT_DISPOSITION_CODE, INCIDENT_DATETIME) %>% 
  janitor::clean_names() %>% 
  separate(col = incident_datetime, into = c('date', 'time'), sep = ' ') %>% 
  separate(col = date, into = c("month","day"), sep = '/') %>% 
  mutate(month = factor(month, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"), labels = c("Jan", "Feb", "Mar", "Apr", "May", "June", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"), ordered = TRUE)) %>% 
  mutate(arrival_outcome = ifelse(incident_disposition_code == "83", "dead", "alive"),
         arrival_outcome = recode_factor(arrival_outcome, `0` = "alive", `1` = "dead"),
         initial_severity_level_code = factor(initial_severity_level_code, 
 levels = c("1", "2","3", "4", "5", "6", "7", "8", "9"), ordered = TRUE),
         final_severity_level_code = factor(final_severity_level_code, 
                                            levels = c("1", "2", "3", "4", "5", "6", "7", "8"), ordered = TRUE), 
         held_indicator = recode(held_indicator, "N" = "no", "Y" = "yes")) %>%
  mutate(neighbourhood = recode(zipcode, "10026" = "central harlem", "10027" = "central harlem", "10030" = "central harlem", "10037" = "central harlem", "10039" = "central harlem", "10001" = "chelsea and clinton", "10001" = "chelsea and clinton", "10011" = "chelsea and clinton", "10018" = "chelsea and clinton", "10019" = "chelsea and clinton", "10020" = "chelsea and clinton", "10036" = "chelsea and clinton",  "10029" = "east harlem", "10035" = "east harlem", "10010" = "gramercy park and murray hill", "10016" = "gramercy park and murray hill", "10017" = "gramercy park and murray hill", "10022" = "gramercy park and murray hill", "10012" = "greenwich village and soho", "10013" = "greenwich village and soho", "10014" = "greenwich village and soho", "10004" = "lower manhattan", "10005" = "lower manhattan", "10006" = "lower manhattan", "10007" = "lower manhattan", "10038" = "lower manhattan", "10280" = "lower manhattan", "10002" = "lower east side", "10003" = "lower east side", "10009" = "lower east side", "10021" = "upper east side", "10028" = "upper east side", "10044" = "upper east side", "10065" = "upper east side", "10075" = "upper east side", "10128" = "upper east side", "10023" = "upper west side", "10024" = "upper west side", "10025" = "upper west side", "10031" = "inwood and washington heights", "10032" = "inwood and washington heights", "10033" = "inwood and washington heights", "10034" = "inwood and washington heights", "10040" = "inwood and washington heights" )
  ) %>%  
  drop_na(neighbourhood) %>% 
    select(-incident_disposition_code) 
```
```{r, warning = FALSE, echo = FALSE}
ems_data_clean %>% 
  select(arrival_outcome, month) %>% 
  group_by(month, arrival_outcome) %>%
  drop_na(arrival_outcome) %>% 
  summarise_(n = ~n()) %>%
  filter(arrival_outcome == "dead") %>% 
  plot_ly(y = ~n, x = ~month, type = "bar",
          color = ~month, alpha = 0.7) %>% 
  layout(title = "Figure 1. Death Count on Arrival for each Month in 2018",
    xaxis = list(title = "Month"),
         yaxis = list(title = 'Dead on Arrival (n)'),
    showlegend = FALSE)
ems_data_clean %>% 
  select(arrival_outcome, neighbourhood) %>% 
  group_by(neighbourhood, arrival_outcome) %>%
  drop_na(arrival_outcome) %>% 
  summarise_(n = ~n()) %>%
  filter(arrival_outcome == "dead") %>% 
  plot_ly(y = ~n, x = ~neighbourhood, type = "bar",
          color = ~neighbourhood) %>% 
  layout(title = "Figure 2. Death Count on Arrival by NYC Neighborhood",
    xaxis = list(title = "Neighborhood"),
         yaxis = list(title = 'Dead on Arrival (n)'),
    showlegend = FALSE)
deaths_scatter = ems_data_clean %>%
select(arrival_outcome, neighbourhood, month) %>%
group_by(arrival_outcome, neighbourhood, month) %>% 
drop_na(arrival_outcome) %>% 
summarise_(n_death = ~n()) %>%
filter(arrival_outcome %in% "dead") %>%
plot_ly(x = ~month, y = ~n_death, color = ~neighbourhood,type = "scatter", mode = "lines+markers", alpha = 0.5) %>%
  layout(title = 'Figure 3. Death Count on Arrival by NYC Neighborhood Over Time, 2018',
    xaxis = list(title = 'Month'),
         yaxis = list(title = 'Dead on Arrival (n)'),
         legend = list(x = 100, y = 0.5))
deaths_scatter
ems_data_clean %>% 
  select(arrival_outcome, incident_travel_tm_seconds_qy) %>% 
  group_by(arrival_outcome) %>% 
  drop_na(arrival_outcome) %>%
  summarize(mean_time = mean(incident_travel_tm_seconds_qy))
time_boro = ems_data_clean %>%
  select(neighbourhood, arrival_outcome, incident_travel_tm_seconds_qy) %>%
  group_by(arrival_outcome, neighbourhood) %>%
  drop_na(arrival_outcome) %>%
  summarize(mean_time = mean(incident_travel_tm_seconds_qy)) %>% 
  plot_ly(x = ~neighbourhood, y = ~mean_time/60, color = ~arrival_outcome, group = ~arrival_outcome, type = "bar", colors = "Set2") %>% 
  layout(title = "Figure 4. Mean Travel Time by NYC Neighborhood",
    xaxis = list(title = "Neighborhood"),
         yaxis = list(title = 'Mean EMS Travel Time (min)'),
    showlegend = TRUE)
time_boro
ems_data_clean %>%
  select(neighbourhood, initial_severity_level_code, incident_travel_tm_seconds_qy) %>%
  group_by(neighbourhood, initial_severity_level_code) %>% 
  summarize(mean_time = mean(incident_travel_tm_seconds_qy)) %>%
  plot_ly(x = ~initial_severity_level_code, y = ~mean_time/60, group = ~neighbourhood, color = ~neighbourhood, type = "scatter", mode = "lines+markers", alpha = 0.5) %>% 
  layout(title = "Figure 5. Mean Travel Time by Severity for each Neighborhood",
    xaxis = list(title = "Severity Code"),
         yaxis = list(title = "Mean Travel Time (min)"),
         legend = list(x = 100, y = 0.5))
```

#

## Statistical analysis

In our statistical analysis, we aimed to answer the following questions:

1. **Is response time significantly associated with clinical outcome regarding survival?** To address this, we ran a two sample t-test comparing mean response times. 

2. **Does response time vary significantly between neighborhoods?** To address this, we performed an Analysis of Variance (ANOVA) test to determine whether mean travel time was statistically significantly different between neighborhoods. 

3. **After controlling for severity of the emergency and neighborhood, is response time associated with clinical outcome?** To address this, we performed a logistic regression, including the two possible confounders of severity and neighborhood in the model. 


#

## Results & Discussion

*Response time and survival*

The results of our t-test, shown below, indicate that there is a statistically significant difference (p < 0.0001) in the response time for those who are found alive upon arrival EMS Response Teams and those who are not. The mean response time for those who were found deceased upon arrival was 4.9 minutes, almost half of the 8.8 minutes for those found alive upon arrival.

This indicates that the emergency situation and its repercussions may have an effect on the response time of EMS Response teams. Thus, if an EMS responder believes an individual may die in the incident, they will respond in a quicker time.

```{r warning=FALSE, echo = FALSE}
t.test(ems_data_clean$incident_travel_tm_seconds_qy/60 ~ ems_data_clean$arrival_outcome, na.rm = FALSE)
```


*Response time and neighborhood*

The results of our ANOVA test indicate that there is a statistically significant difference between mean response time and neighborhood. This ANOVA does not indicate which neighborhoods are different from each other, but that at least one of them has a statistically different mean response time. Figure 4 above displays the mean response times by neighborhood. 

```{r warning=FALSE, echo = FALSE}
res.aov = aov(incident_travel_tm_seconds_qy/60 ~ neighbourhood, data = ems_data_clean)
summary(res.aov) 
```

*Response time and survival, controlling for severity and neighborhood*

```{r include= FALSE, warning=FALSE, echo = FALSE}
ems_data_clean %>%  
    mutate(
    initial_severity_level_code = fct_relevel(initial_severity_level_code, "1")) %>%
    select(incident_travel_tm_seconds_qy, initial_severity_level_code, neighbourhood)
```

We perform a logistic regression model to predict the odds of being dead on arrival. 
Our initial model also included month and severity code as k-level categorical predictors. However, since we did not find any statistically significant relationships between month, severity code, and arrival outcome, we decided to exclude the variables from the model. 

The final model includes travel time in seconds and neighborhood, with Central Harlem as the reference group because it had a death count that was closest to the mean. 

Our results indicate that travel time does not predict the outcome on arrival (OR = 0.996), when adjusting for neighborhood. We do see that there are significant associations between neighborhood and arrival outcome. 

The Upper East Side and Upper West Side show the highest odds of being dead arrival (OR = 1.638 and 1.905). We hypothsize that this might be due to an older population residing in these neighborhoods. These results were statistically significant (p-value < 0.0001) at the 5% level of significance. 

We also find that Lower Manhattan and Chelsea have the lowest odds of being dead on arrival (OR = 0.39 and 0.44). These results were statistically significant (p-value < 0.0001) at the 5% level of significance. This could be due to a younger population living in these areas and also quicker access to hospital services. 

We can conclude that survival odds decrease as we move to the lower parts on Manhattan and are the highest in the Upper East and West Sides. 


#

## DISCUSSION
From our results, we also found that the mean travel time across the 10 NYC neighborhoods varies. EMS responders respond slowest to individuals living in East Harlem, Central Harlem, and Inwood/Washington Heights. This may be due to neighborhood proximity to healthcare systems and/or congestion and commuter traffic within neighborhoods. 

Additionally, based on our analyses, we found that the Upper West Side had the highest number of deaths upon arrival. This was interesting to us, as we initially predicted lower-income neighborhoods would have higher deaths upon arrival. It may possible that the Upper West Side may have the highest number of deaths due to the older average age of the population. We suggest further research into analyzing the death count per neighborhood, especially comparing the count across past years. The odds of death upon arrival was significantly higher in the Upper West Side and Upper East Side. 

We also found that the death count on arrival varies by month. The highest death counts are in January and July, which suggests extreme weather could be affecting the time it takes EMS responders to reach the incident or more individuals may be dying from deaths associated with extreme weather. 

Based on our results, we believe that while improving EMS time is ideal, given limited budgeting, we have identified certain neighborhoods that should be focused on as well as certain time frames throughout the year that should be the focus of further improvement. Further research can look at variables such as age and traffic details.


## Strengths & Limitations

Our study was able to efficiently and clearly assess the relationship between response time, survival, and Manhattan neighborhoods. 

One of the biggest limitations of our study is that the data did not contain a variable indicating the type of response unit sent. The FDNY is in command of many different response units, which all correspond to different types of emergencies.

Additionally, the data is only recorded once the incident is created in the FDNY system. If you’ve ever called 911 (hopefully you haven’t had to), you’ll know that the first person you talk to is not a police officer, emergency medical technician, or fire fighter. It’s a 911 Call Taker, as shown by this schematic provided by [NYC Analytics](http://home2.nyc.gov/html/911reporting/html/anatomy/call.shtml). These call takers ask tons of questions, the most notable being “What’s your emergency” to assess the appropriate emergency response team for the situation. A police officer may not be as helpful as a firefighter for a building fire.

After the dispatcher determines the best response team for the emergency, they will either put you directly in contact with someone from say, the fire department or police, or they will pass the information along to them while they stay on the line with you. (In the NYC Analytics [schematic](http://home2.nyc.gov/html/911reporting/html/anatomy/call.shtml), this is shown under FDNY EMS VESTA in Orange as 911 Call Conference). Then, either the caller or call taker may have to relay the pertinent information to the response team dispatcher.

Our data is collected only after the first 911 call taker has passed the call along to the FDNY EMS (Incident creation in the [schematic](http://home2.nyc.gov/html/911reporting/html/anatomy/call.shtml)). Thus, it does not include the amount of time the caller spent explaining their emergency to the dispatcher before being transferred to the FDNY. In many cases, this can be a lengthy process, but from our data it is impossible to tell. Additionally, it is unclear from the data whether, from the NYC Analytics schematic, whether our incident creation is at the general incident creation between “Call Taking” and “Skip to Dispatch” in yellow at the top, or the “EMS Incident Creation” between “911 Call Conference” and “Dispatch Queue” in orange at the bottom of the schematic.

911 calls are emergencies, so time is crucial. The time before the incident is created could influence not only the resulting morbidity or mortality of those involved in the emergencies, it could also influence the amount of time the FDNY spends on the phone before a unit is dispatched. For example, if the 911 caller is in some way incapacitated, the initial call taker may have difficulties assessing the emergency and identifying the appropriate response team. The FDNY EMS dispatcher may also then have difficulties assessing what type of unit to send, or even where to send the unit.

As a result, the data provide indications of the efficiency of the EMS Response Teams in each neighborhood, but not necessarily the safety or other inherent traits thereof. 

Another limitation to our data is that it was only collected from 2018. Amid political unrest regarding our current President and his legal affairs, Supreme Court Judge pick, and immigration policies; unhappiness with NYC's current mayor; and mass shootings rocking the nation, this year could have seen an incre`ase in violent crime or substance abuse, skewing the results. Conversely, compared to the election year of 2016, crime and substance abuse could have dropped significantly. We cannot extrapolate our findings to years past or present.

#

## Conclusion

The distributions of those deceased upon arrival of EMS Response Teams and travel time may be affected by a number of external factors. Future research comparing political climate, extreme weather, or local cultural events to multiple years of data may provide more insight into the reasoning for the variation seen in our data.



